Blind Search for 21-cm Absorption Systems in New Generation Chinese Radio Telescopes

### 2017-03-15 ms_21cm.tex
v3

\abstract{
Neutral hydrogen clouds are known to exist in the Universe, however their spatial
distributions and physical properties are poorly understood. Such missing information
can be studied by the new generation Chinese radio telescopes through a blind searching
of 21-cm absorption systems. We forecast the capabilities of surveys of 21-cm absorption
systems by two representative radio telescopes in China --
Five-hundred-meter Aperture Spherical radio Telescope (FAST) and
Tianlai 21-cm cosmology experiment (Tianlai).
Facilitated by either the high sensitivity (FAST)
or the wide field of view (Tianlai) of these telescopes,
\tcr{[more than a thousand for FAST (and Tianlai)] (to be determined by final calculations)} 21-cm absorption systems can be discovered in a few years,
representing orders of magnitude improvement over the cumulative discoveries
in the past half a century. 
\keywords{}
}

@@

\section{Sensitivity}\label{sec.sensitivity}
To find possible 21-cm absorption systems, radio telescopes
could be devised to scan the radio sources in
NRAO VLA Sky Survey (NVSS)\footnote{http://www.cv.nrao.edu/nvss/}.
If we assume the distributions of radio sources $n_R$ and
{\HI} clouds $n_{\rm \HI}$ are uncorrelated, the redshift
distribution of the 21-cm absorption systems $n_{\rm system}$ would be
\tcr{(still need a volume filling factor $K(z)$?)}
\begin{equation}\label{eq.n_system}
    n_{\rm system}(z)=\tcr{n_{\rm \HI}(z)}\int_z^\infty n_R(z')\diff z'
\end{equation}
\tcr{(In my definition,
$n_{\rm \HI}$ is the cloud regardless observable or not, 
and $n_{\rm system}$ is the absorber, associated with a background source, absorbing the
source flux on the line of sight to Earth.)}
regardless of the observation sensitivity,
where $n_R(z)$ is given by equation (26) of \cite{2010A&ARv..18....1D}.
Spacial distribution of {\HI} is poorly understood and is to be
studied in upcoming surveys.
Recent studies show that the number density of
absorbers to be $\diff \tcr{n_{\rm \HI}}/\diff z= 0.045\pm0.006$
per line of sight at low and medium redshift
\citep{2005ARA&A..43..861W,2007ASSP....3..501Z}, so we apply this value
in the following forcast.
Observation sensitivity depends on the background fluxes and foreground
absorptions (current section), as well as telescope configurations and
observation strategies (section \ref{sec.FAST},\ref{sec.Tianlai}).

NVSS contains 2 million sources (with declination $\delta >-40^\circ$,
cover 82\% of the sky) stronger than 2.5
mJy at $\nu=$1.4 GHz \citep{1998AJ....115.1693C} and their flux
distribution $n_R(F)$ is given by \cite{1984ApJ...287..461C}.
Redshifted {\HI} clouds absorb lower frequency bands where
the radio sources are typically brighter by $\nu^{-0.7}$
\citep{1998AJ....115.1693C}, or $(1+z_{\rm \HI})^{0.7}$.
The error of the measurement $\Delta F$ is given by\footnote{https://science.nrao.edu/facilities/vla/docs/manuals/oss/performance/sensitivity} \tcr{[confirm this reference]}
\begin{equation}\label{eq.delta_F}
    \tcr{\Delta F=T_{\rm sys}A_{\rm eff}^{-1}/\sqrt{n_{\rm pol}\Delta\nu\Delta t}},
\end{equation}
where
\tcr{$T_{\rm sys}$ and $A_{\rm eff}$ are system temperature and effective receiving area,}
%SEFD \tcr{(or $T_{\rm sys}$ TBD)} is the system equivalent flux density,
$n_{\rm pol}$
is the number of polarizations, $\Delta\nu$ is the line width and
$\Delta t$ is the integration time. $\Delta\nu=(u_{\rm width}/c)\nu_{\rm obs}$,
where $u_{\rm width}$ is the equivalent line width of the {\HI}
absorber. The properties of 21-cm absorbers are primarily derived from
followup studies of optical absorbers. The discovery rate from the
cross correlation is a lower bound on the expect number of absorbers,
since high column density systems in the CNM may systematically
obscure potential background optical sources \citep{2014PhRvL.113d1303Y}. There are
only 3 blind radio detections, and a survey may discover more systems
which are optically obscured. They would likely be cold and at high
column density. For sensitivity purposes, we treat all sources as
$u_{\rm width}=2\ {\rm km\ s}^{-1}$
\citep{1982ApJ...259..495W,2005ARA&A..43..861W}.
$\nu_{\rm obs}$ is the observation frequency. For redshifted
21-cm absorption systems, $\nu_{\rm obs}=1420\ {\rm MHz}/(1+z_{\rm \HI})$.
The integration time $\Delta t=(n_{\rm obs}\times 24\ {\rm h})\tau$,
where $n_{\rm obs}$ is the number of days an object is scanned,
and $\tau=\lambda_{\rm obs}/2\pi D\cos\delta$ is the fractional time the object
transits the FoV ($\lambda_{\rm obs}\ll D$ or $\delta\nrightarrow\pi/2$).
For 21-cm absorption systems $\lambda_{\rm obs}=21\ {\rm cm}\times (1+z_{\rm obs})$.
$\Delta F$ is redshift- (frequency-) independent, as $(1+z_{\rm obs})$ terms
from integration time and line width cancel out
in the equation (\ref{eq.delta_F}).

@@

FAST has a primary dish of 500 meter in diameter. The effective
antenna aperture of FAST is 300 m, in diameter for any zenith angle
up to $26^\circ$ and decreases to 200 m for a maximum zenith angle 
of $40^\circ$ \citep{2016RaSc...51.1060L}.
%This enables its zenith angle up to $40^\circ$ and it can scan objects
%in declination range $-14^\circ 12'<\delta<65^\circ 48'$.
The 19-beam feed-horn array at L band will be the primary survey
instrument for FAST. For a single day it can scan total of \tcr{$0.5^\circ$ (TBD)}
in declination $\delta$. FAST survey strategy scans each object
only once ($n_{\rm obs}=1$), and with its high sensitivity,
even faint sources are identified in a single day. 
We assume a one-month survey by FAST
around the celestial equator for simplicity. FAST observes
1.02 to 1.42 GHz (redshift $z_{\rm \HI}<0.39$), so objects on celestial equator
have average transit time $\Delta t \simeq \tcr{12}\ {\rm sec.}$
Taking $\Delta\nu\simeq$ 9 kHz, $n_{\rm pol}=2$ (dual polarized system),
\tcr{$T_{\rm sys}\sim 20$ K and $A_{\rm eff}\sim 7\times 10^4$ m$^2$,
(how do we get ${\rm SEFD}=1$ Jy for FAST???? In Li \& Pan (2016), I'm also
confused how to get $A/T=2000$ m/K from these knwon numbers.)},
we have $\Delta F \simeq\ 2.3\ {\rm mJy}$.
Taking into account the $\nu^{-0.7}$ flux boost at lower
frequencies, $rF>10\Delta F$ requires $F\gtrsim 0.11\ {\rm Jy}$ on 1.4 GHz.

@@

Because Tianlai locates at latitude
$\phi_{\rm site}=+44^\circ$ and its cylinders are fixed, only
objects at the zenith ($\delta=\phi_{\rm cite}=44^\circ$) fully utilize
$A_{\rm eff}$. On the other hand, higher declination objects
have longer integration time per day. Thus, $\Delta F$ is
a function of $\delta$, 
\begin{equation}\label{eq.delta_F_delta}
    \Delta F(\delta) = \frac{\tcr{T_{\rm sys}A_{\rm eff}^{-1}}}{\cos(\delta-\phi_{\rm site})}
    {\sqrt{\frac{\cos\delta}{n_{\rm pol}\Delta\nu\Delta t(\delta=0)}}},
\end{equation}
where $\Delta t(\delta=0)\simeq 1.17\times 10^5\ {\rm sec}$ is the
total integration time (1-year survey) for objects on celestial equator,
for Tianlai frequency range 800 to 900 MHz ($0.58<z<0.78$).
The effective $N_R$ is given by
\begin{equation}\label{eq.N_R_eff}
    N_R=\int_0^{\pi/2}2\pi\cos\delta
    \left(\int_{10\Delta F(\delta)/r}^{+\infty}
    n_R(F')\diff F'\right)
    \diff\delta.
\end{equation}
Note that the inner flux integration converges as
$F'\rightarrow+\infty$ because $n_R(F)\sim F^{-2.5}$.
Additionally, even if $\tau=\lambda_{\rm obs}/2\pi D\cos\delta$
fails at $\delta\rightarrow\pi/2$ (the north pole
$\tau(\delta=\pi/2)=1$ is always in FoV),
the outer declination integration also converges
as $\delta\rightarrow\pi/2$, because the pole area is
tiny and neglectable.
For FAST, surveys at higher declinations would as well
use equation (\ref{eq.delta_F_delta},\ref{eq.N_R_eff}),
however $\cos(\delta-\phi_{\rm site})$ term affecting
SEFD should be neglected.

>>>>

\abstract{
Neutral hydrogen clouds are known to exist in the Universe, however their spatial
distributions and physical properties are poorly understood. Such missing information
can be studied by the new generation Chinese radio telescopes through a blind searching
of 21-cm absorption systems. We forecast the capabilities of surveys of 21-cm absorption
systems by two representative radio telescopes in China --
Five-hundred-meter Aperture Spherical radio Telescope (FAST) and
Tianlai 21-cm cosmology experiment (Tianlai).
Facilitated by either the high sensitivity (FAST)
or the wide field of view (Tianlai) of these telescopes,
more than a thousand 21-cm absorption systems can be discovered in a few years,
representing orders of magnitude improvement over the cumulative discoveries
in the past half a century. 
\keywords{}
}

@@

\section{Sensitivity}\label{sec.sensitivity}
To find possible 21-cm absorption systems, radio telescopes
could be devised to scan the radio sources in
NRAO VLA Sky Survey (NVSS)\footnote{http://www.cv.nrao.edu/nvss/}.
\tcr{If we assume the distributions of radio sources and
{\HI} clouds are uncorrelated over the sky}, the redshift
distribution of the 21-cm absorption systems $n_{\rm system}$ would be
\begin{equation}\label{eq.n_system}
    n_{\rm system}(z)=n_{\rm \HI}^*(z)\int_z^\infty n_R(z')\diff z',
\end{equation}
\tcr{where $n_R$ is the redshift distribution of radio sources over the sky
(given by equation (26) of \cite{2010A&ARv..18....1D}), and
$n_{\rm \HI}^*(z)$ is the number of occupation of {\HI} clouds
per any given line of sight. Note that, unlike $n_R$ and $n_{\rm system}$,
 $n_{\rm \HI}^*(z)$ is not an
integration of volume so we do not need a volume filling factor in
equation \ref{eq.n_system}.}
However, this spacial distribution of {\HI} is poorly understood and is to be
studied in upcoming surveys. Recent studies show that the number density of
absorbers to be $\diff n_{\rm \HI}/\diff z=n_{\rm \HI}^*(z)= 0.045\pm0.006$
per line of sight at low and medium redshift
\citep{2005ARA&A..43..861W,2007ASSP....3..501Z}, so we apply this value
in the following forcast. \tcr{Equation \ref{eq.n_system} gives the
{\HI} cloud occupation regardless of observabilty, based on which
we calculate the detections considering sensitivities.}

Observation sensitivity depends on the background fluxes and foreground
absorptions (current section), as well as telescope configurations and
observation strategies (section \ref{sec.FAST},\ref{sec.Tianlai}).

NVSS contains 2 million sources (with declination $\delta >-40^\circ$,
cover 82\% of the sky) stronger than 2.5
mJy at $\nu=$1.4 GHz \citep{1998AJ....115.1693C} and their flux
distribution $n_R(F)$ is given by \cite{1984ApJ...287..461C}.
\tcr{The observed flux is already lowered by the 21cm absorption if there is any on
the line of sight, so the actual signal to noise of the absorption line
is higher. Due to the limited knowledge of absorption system properties
we ignore this factor for now.}
Redshifted {\HI} clouds absorb lower frequency bands where
the radio sources are typically brighter by $\nu^{-0.7}$
\citep{1998AJ....115.1693C}, or $(1+z_{\rm \HI})^{0.7}$.
The error of the measurement $\Delta F$ is given by\footnote{https://science.nrao.edu/facilities/vla/docs/manuals/oss/performance/sensitivity}
\begin{equation}\label{eq.delta_F}
    \Delta F=T_{\rm sys}A_{\rm eff}^{-1}/\sqrt{n_{\rm pol}\Delta\nu\Delta t},
\end{equation}
where
\tcr{$T_{\rm sys}$ and $A_{\rm eff}$ are system temperature and effective receiving area,
and $T_{\rm sys}A_{\rm eff}^{-1}$ is just the system equivalent flux density (SEFD).}
%SEFD \tcr{(or $T_{\rm sys}$ TBD)} is the system equivalent flux density,
$n_{\rm pol}$
is the number of polarizations, $\Delta\nu$ is the line width and
$\Delta t$ is the integration time. $\Delta\nu=(u_{\rm width}/c)\nu_{\rm obs}$,
where $u_{\rm width}$ is the equivalent line width of the {\HI}
absorber. The properties of 21-cm absorbers are primarily derived from
followup studies of optical absorbers. The discovery rate from the
cross correlation is a lower bound on the expect number of absorbers,
since high column density systems in the CNM may systematically
obscure potential background optical sources \citep{2014PhRvL.113d1303Y}. There are
only 3 blind radio detections, and a survey may discover more systems
which are optically obscured. They would likely be cold and at high
column density. For sensitivity purposes, we treat all sources as
$u_{\rm width}=2\ {\rm km\ s}^{-1}$
\citep{1982ApJ...259..495W,2005ARA&A..43..861W}.
$\nu_{\rm obs}$ is the observation frequency. For redshifted
21-cm absorption systems, $\nu_{\rm obs}=1420\ {\rm MHz}/(1+z_{\rm \HI})$.
The integration time $\Delta t=(n_{\rm obs}\times 24\ {\rm h})\tau$,
where $n_{\rm obs}$ is the number of days an object is scanned,
and $\tau=\lambda_{\rm obs}/2\pi D\cos\delta$ is the fractional time the object
transits the FoV ($\lambda_{\rm obs}\ll D$ or $\delta\nrightarrow\pi/2$).
For 21-cm absorption systems $\lambda_{\rm obs}=21\ {\rm cm}\times (1+z_{\rm obs})$.
$\Delta F$ is redshift- (frequency-) independent, as $(1+z_{\rm obs})$ terms
from integration time and line width cancel out
in the equation (\ref{eq.delta_F}).

@@

FAST has a primary dish of 500 meter in diameter. The effective
antenna aperture of FAST is 300 m, in diameter for any zenith angle
up to $26^\circ$ and decreases to 200 m for a maximum zenith angle 
of $40^\circ$ \citep{2016RaSc...51.1060L}.
%This enables its zenith angle up to $40^\circ$ and it can scan objects
%in declination range $-14^\circ 12'<\delta<65^\circ 48'$.
The 19-beam feed-horn array at L band will be the primary survey
instrument for FAST. For a single day it can scan total of $0.5^\circ$
in declination $\delta$. FAST survey strategy scans each object
only once ($n_{\rm obs}=1$), and with its high sensitivity,
even faint sources are identified in a single day. 
We assume a one-month survey by FAST
around the celestial equator for simplicity. FAST observes
1.02 to 1.42 GHz (redshift $z_{\rm \HI}<0.39$), so objects on celestial equator
have average transit time $\Delta t \simeq \tcr{12}\ {\rm sec.}$
Taking $\Delta\nu\simeq$ 9 kHz, $n_{\rm pol}=2$ (dual polarized system),
\tcr{$T_{\rm sys}\sim 20$ K and $A_{\rm eff}\sim 7\times 10^4$ m$^2$,
we get ${\rm SEFD}\sim 1$ Jy for FAST},
then we have $\Delta F \simeq\ 2.3\ {\rm mJy}$.
Taking into account the $\nu^{-0.7}$ flux boost at lower
frequencies, $rF>10\Delta F$ requires $F\gtrsim 0.11\ {\rm Jy}$ on 1.4 GHz.

@@

Because Tianlai locates at latitude
$\phi_{\rm site}=+44^\circ$ and its cylinders are fixed, only
objects at the zenith ($\delta=\phi_{\rm cite}=44^\circ$) fully utilize
$A_{\rm eff}$. On the other hand, higher declination objects
have longer integration time per day. Thus, $\Delta F$ is
a function of $\delta$, 
\begin{equation}\label{eq.delta_F_delta}
    \Delta F(\delta) = \frac{T_{\rm sys}A_{\rm eff}^{-1}}{\cos(\delta-\phi_{\rm site})}
    {\sqrt{\frac{\cos\delta}{n_{\rm pol}\Delta\nu\Delta t(\delta=0)}}},
\end{equation}
where $\Delta t(\delta=0)\simeq 1.17\times 10^5\ {\rm sec}$ is the
total integration time (1-year survey) for objects on celestial equator,
for Tianlai frequency range 800 to 900 MHz ($0.58<z<0.78$).
The effective $N_R$ is given by
\begin{equation}\label{eq.N_R_eff}
    N_R=\int_0^{\pi/2}2\pi\cos\delta
    \left(\int_{10\Delta F(\delta)/r}^{+\infty}
    n_R(F')\diff F'\right)
    \diff\delta.
\end{equation}
Note that the inner flux integration converges as
$F'\rightarrow+\infty$ because $n_R(F)\sim F^{-2.5}$.
Additionally, even if $\tau=\lambda_{\rm obs}/2\pi D\cos\delta$
fails at $\delta\rightarrow\pi/2$ (the north pole
$\tau(\delta=\pi/2)=1$ is always in FoV),
the outer declination integration also converges
as $\delta\rightarrow\pi/2$, because the pole area is
tiny and neglectable.
For FAST, surveys at higher declinations would as well
use equation (\ref{eq.delta_F_delta},\ref{eq.N_R_eff}),
however $\cos(\delta-\phi_{\rm site})$ term affecting
SEFD should be neglected.
